---
title: 'Reproducible Research: Peer Assessment 2'
author: "AWP"
date: "Monday, July 20, 2015"
output: html_document
---

##Synopsis


##Data Processing
```{r download_data_and_load_data}

library(dplyr)
#download data if not already downloaded
if(!file.exists("FStormData.csv.bz2")){
  url<-"http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(url, dest="FStormData.csv.bz2", mode="wb") 
  #read in the data from the compresed file
  
}
#We save the raw data to disk as rds, check if it exists, if not create it.
if (!file.exists("raw_data.rds")){
   rawData <- read.csv("FStormData.csv.bz2")
  
  #Convert EVTTYPT to all upper-case and add a new column called EVENTTYPE
  rawData$EVENTTYPE<-toupper(rawData$EVTYPE)
  
  #Narrow our focus
  rawNarrowed<-subset(rawData,select=c(EVENTTYPE,BGN_DATE,MAG,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP,WFO))
  
  #save off the data to a rds format
  saveRDS(rawNarrowed,"raw_data.rds")
  rawData<-rawNarrowed
  rm(rawNarrowed)
}else{
 
  rawData<-readRDS("raw_data.rds")
}
```

##Analysis
```{r initial_analysis}
#Get counts of complete and incomplete data sets
incompleteRows<-sum(!complete.cases(rawData))
completeRows<-sum(complete.cases(rawData))

names(rawData)

#EVTYPE BGN_DATE
#FATALITIES
#INJURIES
#PROPDMG          PROPDMGEXP        CROPDMG          CROPDMGEXP  

#max_stepsByDay<-aggregate(steps ~ interval+date, data = omittedNA, FUN = mean)
require(plyr); 
require(lubridate)
library(dplyr)
```

The raw data has `r format(incompleteRows,big.mark=",",scientific=FALSE)` incomplete rows and `r completeRows` complete rows. Elminiation of incomplete rows is not possilbe at this point and time.


```{r find_duplicate_eventtypes}

#get all of the Eventtypes
allEventTypes<-subset(rawData,select=c(EVENTTYPE))
#find all eventtypes that are duplicates
duplicateEventTypes<-allEventTypes[duplicated(allEventTypes[,c("EVENTTYPE")]),]
#find unique eventtypes from the duplicate list
uniqueDuplicateEventTypes<-unique(duplicateEventTypes)

outcome<-rawData %>% group_by(EVENTTYPE) %>% summarise_each(funs(sum))
nrow(outcome)

outcome[outcome$EVENTTYPE=="TORNADO",]




```

word

```{r}
names(rawData)
xx<-rawData[rawData$EVENTTYPE=="TORNADO",] 

xx



eventType_dates<-subset(rawData,select=c(EVENTTYPE,BGN_DATE)) 
names(eventType_dates)
cleanEventTypeDate<-eventType_dates[!duplicated(eventType_dates[,c("EVENTTYPE","BGN_DATE")]),]
rm(eventType_dates)

cleanEventTypeDate<-mutate(cleanEventTypeDate, date = parse_date_time(BGN_DATE,"%m/%d/%y %H%M%S"),year = year(date))

#head(cleanEventTypeDate)

head(arrange(cleanEventTypeDate,year,EVENTTYPE))

#head(aggregate(cleanEventTypeDate,by=list(year,EVTYPE),FUN=sum))
library(sqldf)
#sqldf('SELECT A, SUM(B) AS B FROM df GROUP BY A')
#numEventsByYear<-sqldf('SELECT EVTYPE, year,count(*) as cnt FROM cleanEventTypeDate where year>=1950 and year<=1954 GROUP BY year,EVTYPE')
```


```{r ploting, fig.width=10}

numEventsByYear<-cleanEventTypeDate %>% group_by(EVENTTYPE,year) %>% tally()
Eliminate50_54<-sqldf('SELECT EVENTTYPE, year,n FROM numEventsByYear where year>=1950 and year<=1954 ')
Eliminate50_54

aggEventType<-aggregate( numEventsByYear$n ~numEventsByYear$EVENTTYPE ,FUN = sum, na.rm=TRUE)
names(aggEventType)<-c("EVENTTYPE","Num")

#Which event has the most occurences 
top_10<-aggEventType %>%  top_n(n=10) 
OrderedTop10<-top_10[order(-top_10$Num),]




library(ggplot2)

ggplot(data=OrderedTop10, aes(x=EVENTTYPE, y=Num)) +geom_bar(stat="identity")+theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))


OrderedTop10

```



```{r}
numEventsByEventType<-cleanEventTypeDate %>% group_by(EVENTTYPE) %>% tally()

aggEventTypeOnly<-aggregate( numEventsByYear$n ~numEventsByYear$EVENTTYPE ,FUN = sum, na.rm=TRUE)

head(aggEventTypeOnly)

top_10Events<-aggEventTypeOnly %>%  top_n(n=10) 
OrderedTop10Events<-top_10Events[order(-top_10$Num),]

OrderedTop10Events

```




##Results
```{r}
```

    1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

    2. Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.